---
title: "Vignette 1: Tutorial on assignment of Proteins to Subcellular Locations Using protlocassign"
author: "Dirk Moore"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output: 
   word_document:
#  html_document:
     #mathjax:  default
     fig_caption:  true
     toc: true
     #section_numbering: true
     #css: ggsci.css


#output: 
#  word_document:
#    reference_docx: word-styles-reference-03.docx
#output: pdf_document
#fontsize: 10pt

vignette: >
  %\VignetteIndexEntry{protlocassign}
  %\VignetteEngine{knitr::rmarkdown}
#  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff = 150
		),
	options(width = 150),
	eval = TRUE
)
```

## Introduction

Determining the locations of proteins in the cell is an important but complex problem. A frequently employed approach for this involves centrifugation-based methods to partially separate different organelles and other cellular compartments and then to determine the relative distribution of different proteins among the centrifugation fractions. The location of proteins of interest in the cellular compartments are then inferred by comparing their distributions across the centrifugation fractions to the distributions of a set of reference proteins (markers) with known cellular locations.

This package implements a subcellular protein assignment procedure as “constrained proportional assignment”, or CPA (Jadot et al, 2016). The basic concepts involved in CPA and analysis of subcellular proteomics data are described in the main paper. These vignettes serve as a basis for implementing CPA and various utilities for both experienced and beginner R users.

There are two fundamental inputs for CPA.  The first is a file with information regarding the distribution of each protein or other species of interest to be analyzed across centrifugation fractions. Each row has a name that serves as a unique identifier, which can be a protein name, protein group, gene name, accession number, or peptide sequence. For each protein identifier, there are data that reflects its distribution among the centrifugal fractions, which represent a profile. The second is a file containing a curated list of marker proteins with defined subcellular locations. As an initial step, the package uses the markers to compute profiles for individual cellular compartments. Then, for each protein, it finds the best match for its profile using a linear combination of compartment profiles. The relative weights of the linear combination in principle reflect the relative abundance of a given protein among different compartments. The CPA method, unlike other previously described methods, can thus account for proteins that have multiple residences, and estimate the relative proportion among these residences. 

We illustrate with an example. Tannous et al. (2020) presented abundance levels of proteins among six fractions from a differential centrifugation experiment (N, M, L1, L2, P, and S), and three fractions from a Nycodenz density gradient separation of the differential fraction L1 (Nyc1, Nyc2, and Nyc3). Eight subcellular compartments were considered: nucleus (Nuc), mitochondria (Mito), lysosomes (Lyso), peroxisomes (Perox), endoplasmic reticulum (ER) Golgi apparatus (Golgi), plasma membrane (PM), and cytosol (Cyto). The CPA method assigns each of a large number of proteins to one or more of these compartments, based on profiles from a set of reference proteins. 

## Installing the protlocassign package
Start by installing the devtools package from CRAN, by typing

```
install.packages("devtools")
```

Then install the protlocassign package from the github repository by typing

```
devtools::install_github("mooredf22/protlocassign0p1p1")
```

This will make the programs and datasets available.
Also, several libraries are required; they may be downloaded from CRAN by typing

```
install.packages(c("BB", "pracma", "lme4"))
```
Once you have installed these packages, you do not need to re-install them the next time you start up R.

These are the only packages required for data that are presented as mean levels for a protein. In another vignette there is a description of R functions that work with data at the spectral level; additional packages will be needed to process spectral-level data and produce protein-level data.

## Working with data

To use this package, you will need two data sets. One, referred to here as the protein profile data set, contains rows with a unique identifier followed by data describing its profile across a series of centrifugation fractions. In this vignette, the profiles are specific amounts with sums constrained to 1 (see main Paper) but can be further transformed to improve the quality of the fit (see Vignettes 2 and 3). In addition, there may be two additional values representing the numbers of peptides and spectra that were used to compute the profile.  The other data set consists of a list of reference proteins and their associated known subcellular compartments.


Consider for example the TMT MS2 data from Tannous et al. (2020). To get started, load the package and look at the dimensions of the embedded protein profile data set which shows that it has 7893 rows and 11 columns: 

```{r, echo=TRUE}
library(protlocassign)
dim(protProfileSummaryTMTms2)
```

Now look at the first few rows of the data using the `head` command, rounding to improve legibility. The first nine columns represent the average levels of the different fractions, across spectra. The last two columns, which are optional, give the number of spectra and the number of peptides (sequences) for each protein. (While we use a nested random effects model described in Jadot, 2016 to compute the means across spectra, other methods, including taking a straight average, may be appropriate for other applications.)

```{r, echo=TRUE}
round(head(protProfileSummaryTMTms2), digits=2)
```

The list containing reference proteins (also called marker proteins) that we use is derived from Jadot et al (2016) and examining the dimensions of the data reveals that it contains 39 rows and two columns, the first for the protein names and the second for their respective subcellular compartments.

```{r, echo=TRUE}
dim(refLocProteinsJadot)
```

The data set can be viewed by entering its name as follows:

```{r, echo=TRUE}
refLocProteinsJadot
```

While the `protProfileSummaryTMTms2` and `refLocProteinsJadot`  data sets are included in the package, they were initially read into R from external files and automatically converted to data frames.  This will need to be done for any new data set.  As an example we demonstrate this for a case where one is working with the Windows operating system and that the data sets reside in a directory `C:\temp\myproteindata`. (If one is working with either the linux or Mac OS, appropriate changes to these procedures will be needed to access files.)
First, set the working directory to point there, either by navigating to it in R studio (session menu pane) or by entering:


```
setwd("C:\\temp\\myproteindata")
```

Note that in R each backslash character must be doubled to be interpreted correctly (since otherwise it will be incorrectly interpreted as an escape character). Alternatively, you may use a single forward slash:

```
setwd("C:/temp/myproteindata")
```

Then read in the two data sets (protein profile data and reference protein list), 
which must be in
comma-separated format (`.csv`), with the first row containing column names. 
The option `row.names=1` takes the first column of
the `MyProtProfile.csv` file and uses it as row names for the R file `MyProtProfile`.

```
MyProtProfile <- read.csv(file="MyProtProfile.csv", row.names=1)
```

For the marker protein data file, the first row must contain the two column names (protein unique identifier and compartment designation). Note that the marker proteins must be specified precisely as listed in the protein profile data set.  

```
MyRefLocProteins <- read.csv(file="MyRefLocProteins.csv", row.names=F)
```


## Creating and viewing compartment and marker protein profiles

In order to assign proteins proportionately to their respective compartments, we first use the function `cpaSetup` to obtain profiles for the compartments based on the means of the individual reference proteins that represent each compartment. This function produces a matrix that has one row for each compartment and one column for each fraction.

```{r, echo=TRUE, eval=TRUE}
markerLocRuse <- cpaSetup(protProfileSummary=protProfileSummaryTMTms2, refLocProteins=refLocProteinsJadot, n.channels=9)
```

We display the data, using rounding to improve readability.

```{r, echo=TRUE, eval=TRUE}
round(markerLocRuse, digits=4)
```

To examine the available compartments, view the row names of `markerLocRuse`.

```{r, echo=TRUE, eval=TRUE, fig.show='hold'}
rownames(markerLocRuse)
```

To graphically display a particular compartment profile and its component proteins, use `refProfilePlot`. 
For example, to plot the profile for plasma membrane, use this:

```{r, echo=TRUE, eval=TRUE, fig.show='hold'}

refProfilePlot(refLoc="PM", refLocProteins=refLocProteinsJadot,
                     protProfileSummary=protProfileSummaryTMTms2,
                     markerLocR=markerLocRuse)
```

We see, in the plot title, that this displays the PM compartmental profile (dashed black line) and its five component reference proteins (red lines). To plot, for example, the second PM marker protein, use the option `refProtPlot=2` in the `refProfilePlot` function:

```{r, echo=TRUE, eval=TRUE, fig.show='hold'}

refProfilePlot(refLoc="PM", refLocProteins=refLocProteinsJadot,
                     protProfileSummary=protProfileSummaryTMTms2,
                     markerLocR=markerLocRuse, refProtPlot=2)
```

To show all five of the reference proteins that are used to calculate the PM compartmental profile, use `par(mfrow=c(3,2))` to set up a plot array with 3 rows and 2 columns which will accommodate up to six plots on a single page. Then plot all five of the them one-by-one by looping through the five PM marker proteins by first specifying `for (j in 1:5)` and invoking the option `refProtPlot=j` in the `refProfilePlot` function. (To make them legible, you may first need to open a new window by entering `windows(width=5, height=7)`.) The parameters in `par(mfrow=c(3,2))` and `for (j in 1:5)` can be adjusted to display any number of plots.


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 5, fig.height = 7}

par(mfrow=c(3,2))
for (j in 1:5) {
  refProfilePlot(refLoc="PM", refLocProteins=refLocProteinsJadot,
                     protProfileSummary=protProfileSummaryTMTms2,
                     markerLocR=markerLocRuse, refProtPlot=j)
  }
```

To plot all of the compartment and individual marker protein profiles, we may set up a plot window with four rows and two columns and then loop through the eight subcellular compartments.

```{r, echo=TRUE, eval=TRUE, fig.width = 7, fig.height = 10}
loc.list <- rownames(markerLocRuse)
n.loc <- length(loc.list)
par(mfrow=c(4,2))
for (i in 1:n.loc) {
  refProfilePlot(refLoc=loc.list[i], refLocProteins=refLocProteinsJadot,
                     protProfileSummary=protProfileSummaryTMTms2,
                     markerLocR=markerLocRuse)
  }
```

## Obtaining constrained proportionate assignments of proteins to compartments  

Once we are satisfied with the marker set and compartment profiles, we can run the CPA routine which, for each protein in our data set, apportions its residency among the different specified compartments. This  may take several minutes to complete. 

```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
assignPropsUse <- proLocAll(protProfileSummary=protProfileSummaryTMTms2,
                            markerLocR=markerLocRuse, n.channels=9)
```

We can view the structure of the output data file and see that the row names contain the protein names, and the next 8 columns contain the allocation proportions of each protein to the eight compartments. The last two columns are integers representing the numbers of peptides and spectra, which are carried over from the input `protProfileSummary` profile data, these are not essential for the CPA analysis.

```{r, echo=TRUE, eval=TRUE}
round(head(assignPropsUse), digits=2)
```



We can look at the profile of, for example, the protein "TLN1". We first ensure that the protein is in the dataset:

```{r, echo=TRUE, eval=TRUE}
protIndex("TLN1", protProfileSummary=protProfileSummaryTMTms2)
```

This function also accepts partial matching of the first few letters of a protein. For example, we can find the indices of the proteins starting with "TLN":

```{r, echo=TRUE, eval=TRUE}
protIndex("TLN")
```



Now we plot the results for protein TLN1:

```{r, echo=TRUE, eval=TRUE, fig.width=7, fig.height=9}
protPlotfun(protName="TLN1", protProfileSummary=protProfileSummaryTMTms2, 
n.fractions=9, n.compartments=8,
 Nspectra=T, markerLocR=markerLocRuse, assignPropsMat=assignPropsUse)
```                        

The horizontal axis represents the nine fractions, which are N, M, L1, L2, P, S, Nyc.1, Nyc.2, and Nyc.3. In each of the eight plots, the red line is the average profile of the protein. The dashed yellow-black lines show the expected profile for a protein entirely resident in the respective subcellular location. In this set of plots, we see that the CPA procedure assigns a 59 percent residence proportion to plasma membrane and 35 percent residence to cytosol. Visually, we see that the observed red profile closely matches a mixture of the expected yellow-blacklines. 

The `protPlotfun` function is designed to plot profiles of eight subcellular locations. If a data set has more than eight of these, it will be necessary to modify the code to accommodate the larger number.

## Saving the CPA output
### Data output

To save the results of the CPA procedure, first specify a directory for the output.

```
setwd("C:/temp/myProteinOutput")
```
Next, write out the data as a comma separated file, using `write.csv`:

```
write.csv(assignPropsUse, file="assignPropsUse.txt", row.names=T)
```

This writes the CPA results (which we had previously put into the R dataset `assignPropsUse`), writing row names, and in comma separated format. Ordinarily one would give the file a `.csv` extension so that Excel would automatically open it when clicking on the file name in Windows. Here we do not recommend that because of a naming issue with proteins. Specifically, some proteins (`MARCH9` for example) will be interpreted by Excel as dates, and switched to a date format (`9-Mar` in this case). To prevent that, save the comma-separated value file as a text file, with extension `.txt`. Then, in Windows, open up the Excel program and navigate within Excel to the `assignPropsUse.txt` file and open it. You will then be presented with an Excel import wizard where you specify that columns are separated by commas, and then select the first row (`protName`), and specify that this is a text column. This will prevent Excel from converting protein names to date format.

### Plot output as pdf files

To save the plot of a protein (`TLN1` for example) as a pdf file, we first specify a pdf plot `window`, call the `protPlotfun` function as before, and then close the plot window using `dev.off()` to allow R to complete producing the file:

```
pdf(file="myPlotPDFfile.pdf", width=7, height=10)
protPlotfun(protName="TLN1", protProfileSummary=protProfileSummaryTMTms2, 
    n.fractions=9, n.compartments=8,
    Nspectra=T, markerLocR=markerLocRuse, assignPropsMat=assignPropsUse)
dev.off()
```   

To output plots all of the protein profiles into a single pdf file, one can set up a loop as follows:


```
pdf(file="allPlotsPDFfile.pdf", width=7, height=10)
n.prots <- nrow(assignPropsUse)
for (i in 1:n.prots) {
   protPlotfun(protName=rownames(assignPropsUse)[i],
       protProfileSummary=protProfileSummaryTMTms2, n.fractions=9, n.compartments=8,
       Nspectra=T, markerLocR=markerLocRuse, assignPropsMat=assignPropsUse)
}
dev.off()
```   

This will result in a single file, `allPlotsPDFfile.pdf` with 7893 pages, one for each protein.




## References

Jadot M, Boonen M, Thirion J, Wang N, Xing J, Zhao C, Tannous A, Qian M, Zheng H, Everett JK, Moore DF, Sleat DE, Lobel P (2016) Accounting for protein subcellular localization: a compartmental map of the rat liver proteome. Molecular and Cellular Proteomics 16, 194-212. doi:10.1074/mcp.M116.064527  PMCID: PMC5294208

Tannous A, Boonen M, Zheng H, Zhao C, Germain C, Moore D, Sleat D, Jadot M, Lobel P. (2020) Comparative Analysis of Quantitative Mass Spectrometric Methods for Subcellular Proteomics. Journal of Proteome Research. Journal of Proteome Research 19, 1718-1730.
doi.org/10.1021/acs.jproteome.9b00862


