---
title: "Vignette 5: Calculation of distances between proteins"
author: "Dirk Moore"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output:
  word_document:
  #  html_document:
  #mathjax:  default
  fig_caption:  true
toc: true
#section_numbering: true
#css: ggsci.css
#output:
#  word_document:
#    reference_docx: word-styles-reference-03.docx
#output: pdf_document
#fontsize: 10pt
vignette: >
  %\VignetteIndexEntry{protlocassign}
  %\VignetteEngine{knitr::rmarkdown}
---

  ```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(
    keep.blank.line = TRUE,
    width.cutoff = 150
  ),
  options(width = 150),
  eval = TRUE
)
```


We may find the proteins with profiles nearest to a given protein using the function "nearestProts". Distance is computed as the Euclidean distance between profiles. To use the function, we first use the R function `dist` to create a distance matrix for the proteins in a list of mean profiles, such as `protProfileNSA_AT5tmtMS2`. For clarity of presentation, we rename the embedded data sets to remove experiment-specific labels.

```{r, echo=TRUE, eval=TRUE}
library(protlocassign)
protNSA <- protNSA_AT5tmtMS2
totProt <- totProtAT5
distUse <- dist(protNSA[,1:9], method="euclidean")
```

Then select the protein names:

```{r, echo=TRUE, eval=TRUE}
protsUse <- rownames(protNSA)
```
Finally, provide a protein name. Here, for the protein "CTSD", we find the 10 nearest proteins.

```{r, echo=TRUE, eval=TRUE}
nearestProts(protName="CTSD", n.nearest=10,  distProts=distUse, protNames=protsUse,
             profile=protNSA)
```

Instead of using normalized specific amounts, we may transform them to relative specific amounts:

```{r, echo=TRUE, eval=TRUE}
protProfileLevelsRSA <- RSAfromNSA(NSA=protNSA[,1:9],
                                 NstartMaterialFractions=6, totProt=totProt)
distUseRSA <- dist(protProfileLevelsRSA, method="euclidean")
nearestProts(protName="CTSD", n.nearest=10,  distProts=distUseRSA, protNames=protsUse,
             profile=protProfileLevelsRSA)

```

## References

Jadot M, Boonen M, Thirion J, Wang N, Xing J, Zhao C, Tannous A, Qian M, Zheng H, Everett JK, Moore DF, Sleat DE, Lobel P (2016) Accounting for protein subcellular localization: a compartmental map of the rat liver proteome. Molecular and Cellular Proteomics 16, 194-212. doi:10.1074/mcp.M116.064527  PMCID: PMC5294208

Tannous A, Boonen M, Zheng H, Zhao C, Germain C, Moore D, Sleat D, Jadot M, Lobel P. (2020) Comparative Analysis of Quantitative Mass Spectrometric Methods for Subcellular Proteomics. Journal of Proteome Research. Journal of Proteome Research 19, 1718-1730. doi: 10.1021/acs.jproteome.9b00862

