---
title: 'Vignette 4: More on mixtures and transformations'
author: "Dirk Moore"
date: "`r Sys.Date()`"
output:
  #html_document:
  #  toc: yes
  #  df_print: paged
  word_document:
  fig_caption: yes
toc: yes
#pdf_document:
#  toc: yes
vignette: |
  %\VignetteIndexEntry{protlocassign} %\VignetteEngine{knitr::rmarkdown}
---


  ```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(
    keep.blank.line = TRUE,
    width.cutoff = 150
  ),
  options(width = 150),
  eval = TRUE
)
```
In this vignette we explore briefly the effect of using the normalized specific amounts (NSAs), relative specific amounts (RSAs), and relative amounts (Acup), and their log transformations on CPA.  We also discuss the underlying reasons why the different transformations can yield different results, which may be useful in choosing the appropriate transformation for a particular protein distribution. We illustrate using two mixtures: Cyto with Lyso and Cyto with Nuc. We begin by creating the NSA profiles for the reference compartments, which we will use for further transformations and to create mixtures.

## Applying CPA to relative specific amounts (RSA's), normalized specific amounts (NSA's), or relative amounts (Acup) derived from simulated mixtures of Cyto and Lyso

```{r, echo=TRUE, eval=TRUE}
library(protlocassign)
library(pracma)
refLocationProfiles <- locationProfileSetup(profile=profilesProtAT5tmtMS2, markerList=markerListJadot, numDataCols=9)
AcupMarkers <- relAmtTransform(refLocationProfiles, NstartMaterialFractions=6, totProt=totProtAT5)
```


Now create markers using AcupMarkers Cyto (row 1) and Lyso (row 4) in the standard way, and transform the mixtures to relative specific amounts

```{r, echo=TRUE, eval=TRUE}
i=1
j=4
mixProt1Prot4 <- proteinMix(AcupMarkers, Loc1=i, Loc2=j)
mixProt1Prot4RSA <- RSAfromAcup(Acup=mixProt1Prot4$relAmount, NstartMaterialFractions=6, totProt=totProtAT5)
```

Next, obtain CPA estimates using the RSA-transformed mixtures and RSA-transformed references profiles. We see that the CPA estimates match the proportions we used to create the mixtures:

```{r, echo=TRUE, eval=TRUE}
refLocationProfilesRSA <- RSAfromNSA(NSA=refLocationProfiles, NstartMaterialFractions=6,
        totProt=totProtAT5)
mixProt1Prot4PropRSA <- fitCPA(profile=mixProt1Prot4RSA, 
                               refLocationProfiles=refLocationProfilesRSA, numDataCols=9)
round(mixProt1Prot4PropRSA, digits=3)
```

We may obtain the normalized specific amounts transformation of `mixProtiProtjRSA` by constraining the row sums to 1, as discussed in Vignette 3. We do this as follows:

```{r, echo=TRUE, eval=TRUE}
mixProt1Prot4NSA <- t(apply(mixProt1Prot4RSA,1, function(x) x/sum(x)))
round(mixProt1Prot4NSA, digits=3)
```

Following Vignette 3, we may see what happens if we apply the CPA routine to these normalized specific amounts (NSA's)
```{r, echo=TRUE, eval=TRUE}

mixProt1Prot4PropNSA <- fitCPA(profile=mixProt1Prot4NSA,
                                      refLocationProfiles=refLocationProfiles,
                             numDataCols=9)
round(mixProt1Prot4PropNSA, digits=3)
```

As noted in Vignette 3, the estimates of the proportions deviate somewhat from those used to generate the mixtures.

Now, instead of transforming the Acup mixtures to RSA's or using NSA's, what happens if we just use the Acup mixtures themselves, i.e., the relative amounts? The results show significant departures of the CPA estimates from the proportions used to create the mixtures:

```{r, echo=TRUE, eval=TRUE}
mixProt1Prot4PropAcup <- 
                  fitCPA(profile=mixProt1Prot4$relAmount,
                                   refLocationProfiles=AcupMarkers, numDataCols=9)
round(mixProt1Prot4PropAcup, digits=3)
```

To understand why, here are the AcupMarkers. Note that the three Nyc columns, which are important for classifying lysosomal proteins, are very small, which effectively down-weights their importance in the CPA procedure:

```{r, echo=TRUE, eval=TRUE}
round(AcupMarkers, digits=5)
```

Following are plots of the CPA estimated vs actual mixture proportions for RSA-transformed values, NSA's, and relative (Acup) amounts.

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 4}
par(mfrow=c(1,3))
mixturePlot(mixProtiProtjProp=mixProt1Prot4PropRSA, NstartMaterialFractions=6,
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop, errorReturn = T,
            subTitle="RSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot4PropNSA, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot4$input.prop, errorReturn = T,
            subTitle="NSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot4PropAcup, NstartMaterialFractions=6, 
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop, errorReturn = T,
            subTitle="RelAmt")

```

We may obtain only the areas by using the `mixtureAreaError` function.

```{r, echo=TRUE, eval=TRUE}
mixtureAreaError(mixProtiProtjProp=mixProt1Prot4PropRSA, 
            NstartMaterialFractions=6,
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop)
mixtureAreaError(mixProtiProtjProp=mixProt1Prot4PropNSA, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot4$input.prop)
mixtureAreaError(mixProtiProtjProp=mixProt1Prot4PropAcup, 
            NstartMaterialFractions=6, 
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop)
```

## Applying CPA to log transformations of Cyto and Lyso mixtures

As an alternative to RSA, let us apply a log2 transformation of the Acup values. The resulting CPA estimates are much closer to the true values, but not as close as we had obtained using the RSA transformation:

```{r, echo=TRUE, eval=TRUE}
eps <- 0.001


mixProt1Prot4PropRSALog2 <- fitCPA(profile=log2(mixProt1Prot4RSA + eps), 
                               refLocationProfiles=log2(refLocationProfilesRSA + eps), numDataCols=9)
round(mixProt1Prot4PropRSALog2, digits=3)

mixProt1Prot4PropNSALog2 <- 
            fitCPA(profile=log2(mixProt1Prot4NSA + eps),
                        refLocationProfiles=log2(refLocationProfiles + eps), numDataCols=9)
round(mixProt1Prot4PropNSALog2, digits=3)

mixProt1Prot4PropAcupLog2 <- 
            fitCPA(profile=log2(mixProt1Prot4$relAmount + eps),
                        refLocationProfiles=log2(AcupMarkers + eps) , numDataCols=9)
round(mixProt1Prot4PropAcupLog2, digits=3)

```

Following are plots of the CPA estimated vs actual mixture proportions for simulated proteins where the different profiles (relative specific amounts (RSA's), normalized specific amounts (NSA's), and relative amounts (Acup) are log2-transformed.

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 4}
par(mfrow=c(1,3))
mixturePlot(mixProtiProtjProp=mixProt1Prot4PropRSALog2, NstartMaterialFractions=6,
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop, errorReturn = T,
            subTitle="Log2RSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot4PropNSALog2, NstartMaterialFractions=6, 
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop, errorReturn = T,
            subTitle="Log2NSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot4PropAcupLog2, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot4$input.prop, errorReturn = T,
            subTitle="Log2RelAmt")
```

Here are just the areas:

```{r, echo=TRUE, eval=TRUE}
mixtureAreaError(mixProtiProtjProp=mixProt1Prot4PropRSALog2, 
            NstartMaterialFractions=6,
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop)
mixtureAreaError(mixProtiProtjProp=mixProt1Prot4PropNSALog2, 
            NstartMaterialFractions=6, 
            Loc1=i, Loc2=j, input.prop=mixProt1Prot4$input.prop)
mixtureAreaError(mixProtiProtjProp=mixProt1Prot4PropAcupLog2, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot4$input.prop)
```

## Applying CPA to RSA's, NSA's, or Acup derived from simulated mixtures of Cyto and Nuc

In the previous sections we showed that one can apply CPA to Cyto-Lyso mixtures using a range of transformations, and we saw that the best results were obtained using relative specific values, and the worst results from using relative amounts (Acup transformations). Then we saw that log-transformations of the relative amounts improved the quality of the estimates considerably. 

In this section we consider Cyto and Nuc mixtures, which we generate in the standard way. We shall see that here, Acup-transformed values produce good CPA estimates, just as RSA-transformed values. The reason is that, unlike with Lyso, the Nuc (and Cyto) profiles do not depend on the Nyc portions of the values, and thus the estimated subcellular residence proportions do not suffer from the extremely small Nyc profile values.

Here are the CPA estimates from RSA-transformed profiles:

```{r, echo=TRUE, eval=TRUE}
i=1  # Cyto
j=6  # Nuc
mixProt1Prot6 <- proteinMix(AcupMarkers, Loc1=i, Loc2=j)
mixProt1Prot6RSA <- RSAfromAcup(Acup=mixProt1Prot6$relAmount, 
                                NstartMaterialFractions=6, totProt=totProtAT5)


mixProt1Prot6Prop <- fitCPA(profile=mixProt1Prot6RSA, 
                               refLocationProfiles=refLocationProfilesRSA, numDataCols=9)
round(mixProt1Prot6Prop, digits=3)
```
As noted earlier, the simulated proportions are estimated very accurately.

Following Vignette 3, we may see what happens if we apply the CPA routine to profiles containing  NSA's:

```{r, echo=TRUE, eval=TRUE}
mixProt1Prot6NSA <- t(apply(mixProt1Prot6RSA,1, function(x) x/sum(x)))

mixProt1Prot6PropNSA <- 
                    fitCPA(profile=mixProt1Prot6NSA,
                    refLocationProfiles=refLocationProfiles, numDataCols=9)
round(mixProt1Prot6PropNSA, digits=3)
```

As noted in Vignette 3, the estimates of the proportions deviate somewhat from those used to generate the mixtures.

Now do this using the relative amounts for markers (“Acup markers”) and simulated protein mixtures ($RelAmount)instead of RSA-transformed values.


```{r, echo=TRUE, eval=TRUE}
mixProt1Prot6PropAcup <- fitCPA(profile=mixProt1Prot6$relAmount,
                                   refLocationProfiles=AcupMarkers, numDataCols=9)
```

Note that, unlike with the Cyto-Lyso mixture, the estimates with the Cyto-Nuc mixture are very accurate:

```{r, echo=TRUE, eval=TRUE}
round(mixProt1Prot6PropAcup, digits=3)
```

The CPA for these different transformations can be plotted versus the true proportions as before:

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 4}
par(mfrow=c(1,3))
mixturePlot(mixProtiProtjProp=mixProt1Prot6Prop, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot6$input.prop, errorReturn = T,
            subTitle="RSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot6PropNSA, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot6$input.prop, errorReturn = T,
            subTitle="NSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot6PropAcup, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot6$input.prop, errorReturn = T,
            subTitle="RelAmt")
```


## Applying CPA to log transformations of Cyto and Nuc mixtures

Now try a log2 transformationon different types of simulated protein and marker profiles (RSA's, NSA's, and Acup). Note that the CPA values are different.

```{r, echo=TRUE, eval=TRUE}
eps <- 0.001

mixProt1Prot6PropLog2 <- 
           fitCPA(profile=log2(mixProt1Prot6RSA + eps), 
                      refLocationProfiles=log2(refLocationProfilesRSA + eps), numDataCols=9)
round(mixProt1Prot6PropLog2, digits=3)

mixProt1Prot6PropNSALog2 <- 
           fitCPA(profile=log2(mixProt1Prot6NSA + eps),
                      refLocationProfiles=log2(refLocationProfiles + eps), numDataCols=9)
round(mixProt1Prot6PropNSA, digits=3)


mixProt1Prot6PropAcupLog2 <- 
           fitCPA(profile=log2(mixProt1Prot6$relAmount + eps),
                           refLocationProfiles=log2(AcupMarkers + eps), numDataCols=9)
round(mixProt1Prot6PropAcupLog2, digits=3)
```

Following are plots of the CPA estimated vs actual proportions using the log2-transformed profiles.

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 4}
par(mfrow=c(1,3))
mixturePlot(mixProtiProtjProp=mixProt1Prot6PropLog2, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot6$input.prop, errorReturn = T,
            subTitle="Log2RSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot6PropNSALog2, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot6$input.prop, errorReturn = T,
            subTitle="Log2NSA")
mixturePlot(mixProtiProtjProp=mixProt1Prot6PropAcupLog2, 
            NstartMaterialFractions=6, Loc1=i, Loc2=j,
            input.prop=mixProt1Prot6$input.prop, errorReturn = T,
            subTitle="Log2RelAmt")
```

If desired, we could obtain only the areas using the `mixtureAreaError` function, as we did earlier.

## Transformation fitting error heatmaps

We may visualize the area-based errors for all pairwise mixtures using the `mixtureHeatMap` function, which requires prior installation of the `plot.matrix` R library. For each pair of compartments, this function first creates the mixtures as described earlier using the `Acup` markers. Then it computes the three transformations we have discussed, as well as the log2 transformations of these, and plots them as a 2 by 3 array, with the three transformations (RSA's, NSA's, and Acup) as columns. The first row is the original values, and the second row is a log2 transformation of these values. Colors range from light yellow (for little error) up to red and then to purple (highest error). These 2 by 3 heat maps are arranged in an upper triangular array, with each entry corresponding to a mixture of the row compartment and column compartment.

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 4}
par(mfrow=c(1,1))
mixtureHeatMap(Acup=AcupMarkers, totProt=totProtAT5)
```

The following table summarizes the quantities plotted in each two-by-three rectangle.


```{r, echo=FALSE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 2}
  protAmtList <- c("Relative specific amount\n(RSA)",
                   "Normalized specific amount\n(NSA)",
                   "Relative amount\n(Acup)",
                   "Log2 RSA", "Log2 NSA", "Log2 Acup")
  protAmtMat <- matrix(protAmtList, nrow=2, byrow=T)

  library(grid)
  library(gridExtra)

  x <- c(0,1)
  y <- c(0,1)
  plot(y ~ x, type="n", axes=F, xlab="", ylab="")
  grid.table(protAmtMat)

```

  
This heatmap confirms that relative specific amounts provide the most accurate estimates of the true mixtures, and that relative amounts often (but not always) provide the most inaccurate estimates. The log2-transformed errors (second row of each 2 by 3 heatmap) are never as good as relative specific amounts, but generally improve on estimates based on relative amounts.

## Plotting all pairs of mixtures

We may use the function `mixturePlotPanel` to make these plots for all 8*7/2 = 28 possible pairs of mixtures. The function does this for either RSA-transformed mixtures (`fitType="rsa"`) or mixtures that have been further transformed to normalized specific amounts (`fitType="NSA"`). An option is also available to transform either of these using `y = log2(x + eps)`, where `eps` is a small number; by default, `eps = 0.001`. The option `errorReturn = T` causes the function to return a table of all area-based errors. By summing the third column, we obtain a global measure of error. Because of the large number of pairs, this function must plot to an external pdf file in order to properly format the plots. While these plots cannot be included as part of this vignette, a user may create them using the code below. This code will produce all four combinations of fit type (rsa or specfAmt) and log transformation (True or False). 

First is rsa without and with log-transformed values. The user should set the working directory to the desired location on their device; here we use `C:/temp`:

``` 
setwd("c:\\temp")

fitType <- "rsa"
log2Transf <- F

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtAT5, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])


fitType <- "rsa"
log2Transf <- T

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtAT5, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf, eps=0.001)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])

```

Next are using NSA's, without and with log-transformed values:

``` 

fitType <- "NSA"
log2Transf <- F

pdf(file=paste("CPA assignProts Jadot", fitType=fitType, transf=log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtAT5, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])


fitType <- "NSA"
log2Transf <- T

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtAT5, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf, eps=0.001)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])

```

For completeness we also present plots CPA mixture estimates based on relative amounts and log relative amounts. The use of relative amounts is discussed in detail in Vignette 4.

``` 

fitType <- "relAmt"
log2Transf <- F

pdf(file=paste("CPA assignProts Jadot", fitType=fitType, transf=log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtAT5, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])


fitType <- "relAmt"
log2Transf <- T

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtAT5, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf, eps=0.001)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])

```

## References

Jadot M, Boonen M, Thirion J, Wang N, Xing J, Zhao C, Tannous A, Qian M, Zheng H, Everett JK, Moore DF, Sleat DE, Lobel P (2016) Accounting for protein subcellular localization: a compartmental map of the rat liver proteome. Molecular and Cellular Proteomics 16, 194-212. doi:10.1074/mcp.M116.064527  PMCID: PMC5294208

Tannous A, Boonen M, Zheng H, Zhao C, Germain C, Moore D, Sleat D, Jadot M, Lobel P. (2020) Comparative Analysis of Quantitative Mass Spectrometric Methods for Subcellular Proteomics. Journal of Proteome Research. Journal of Proteome Research 19, 1718-1730. doi: 10.1021/acs.jproteome.9b00862




