---
title: 'Vignette 3: Data Transformations: Notation and Mixtures Using protlocassign'
author: "Dirk Moore"
date: "`r Sys.Date()`"
output:
  #html_document:
  #  toc: yes
  #  df_print: paged
  word_document:
    fig_caption: yes
    toc: yes
  #pdf_document:
  #  toc: yes
vignette: |
  %\VignetteIndexEntry{protlocassign} %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff = 150
		),
	options(width = 150),
	eval = TRUE
)
```

## Adjustment of protein profiles

As explained in the main text and in Vignette 2, there are different ways to transform protein profile data.  Here, we describe a way to explore the effect of using transformed data with CPA.  Briefly, we conduct different data transformations on a set of theoretical proteins that have a range of distributions between two cellular compartments, and then conduct CPA and determine how well it predicts the original distribution.  For our simulations, to create the theoretical proteins, we use data from the experiment from Tannous et al which consists of a TMT-MS2 analysis of six differential fraction (N, M, L1, L2, P, and S) and three fractions from a Nycodenz density gradient separation of the differential fraction L1 (Nyc1, Nyc2, and Nyc3).

In our procedure, we first use the eight compartment profiles generated from the marker protein set to simulate a set of eight theoretical proteins that wholly reside in each of the respective compartments.  We then mix these in defined amounts to simulate proteins that are distributed in varying proportions between two compartments. Note that data from mass spectrometry experiments represent specific amounts ${s_{\alpha ,l}}$ of a protein $\alpha$ in fraction $l$ (e.g. for each fraction, the ratio of the signal in arbitrary units for a given protein $\alpha$ to units of total protein).  For conducting our simulations, we first must transform this data into relative amounts, so that each protein has precisely the same total amount in the initial starting material used for fractionation.


Consider an $n$ by 9 matrix of specific amounts that details the average distribution of $n$ proteins (or other species of interest) among 9 fractions:
$$
\mathbf{S}=\left[ \begin{matrix}
   {{s}_{11}} & {{s}_{12}} & \cdots  & {{s}_{19}}  \\
   {{s}_{21}} & {{s}_{22}} & \cdots  & {{s}_{29}}  \\
   \vdots  & \vdots  & {} & \vdots   \\
   {{s}_{n1}} & {{s}_{n2}} & \cdots  & {{s}_{n9}}  \\
\end{matrix} \right]
$$

Each row $\alpha$ represents a mean profile for a protein or other species of interest $\alpha$, with each profile consisting of $f=9$ fractions. 

In the Vignette 1 we actually used a normalized specific amount $\tilde{s}_{\alpha,l}$  calculated as follows:


$$ {{\tilde{s}}_{\alpha ,l}}=\frac{{{s}_{\alpha ,l}}}{\sum\limits_{j=1}^{f}{{{s}_{\alpha ,j}}}} $$
In some experiments, where bookkeeping is not practicable, these are the only values available for using the CPA procedure. However, with appropriate bookkeeping, the amounts of total protein, $t_l$, are available. We can use these to calculate the amount of protein in arbitrary units in fraction $l$ derived from a set amount of starting material as $a_l = s_l t_l$. Note that as $a_l$ is in arbitrary units one can do the same calculation using either $s_l$ or ${\tilde s_l}$, as later, these will yield the same values when calculating relative amounts (see below).

  To see how this works in `protlocassign`, let us consider the Jadot reference protein profiles

```{r, echo=TRUE, eval=TRUE}
library(protlocassign)
markerLocR <- cpaSetup(protProfileSummary=protProfileSummaryTMTms2, refLocProteins=refLocProteinsJadot, n.channels=9)
round(markerLocR, digits=3)

```


Here, each row represents the profile $\tilde{s}_l$ for a protein resident solely in a particular cellular compartment. The amount of starting material, `tmtMS2totProt`, ($t_l$) in each fraction is as follows:

```{r, echo=TRUE, eval=TRUE}
options(digits=2)
totProtUse <- tmtMS2totProt
totProtUse
```

We denote these values using a vector $\mathbf{t}= ({t_1},{t_2}, \ldots ,{t_9})$. The total amount of starting material ${{t}_{h}}=\sum\limits_{i=1}^{6}{{{t}_{i}}}$ is the sum of the amounts given in the first six (differential) fractions (N, M, L1, L2, P, and S).

```{r, echo=TRUE, eval=TRUE}
sum(totProtUse[1:6])
```

(The last three "Nyc" columns were derived from L1, so we do not include them in the sum.)

As noted above, for protein $\alpha$ in fraction $l$, we have ${{a}_{\alpha ,l}}={{\tilde{s}}_{\alpha ,l}}{{t}_{l}}$. In matrix form,


$\mathbf{A}=\widetilde{\mathbf{S}}\cdot\mathrm{diag}(\mathbf{t})=\left[\begin{matrix}{\widetilde{s}}_{11}t_1&{\widetilde{s}}_{12}t_2&\cdots&{\widetilde{s}}_{19}t_9\\{\widetilde{s}}_{21}t_1&{\widetilde{s}}_{22}t_2&\cdots&{\widetilde{s}}_{29}t_9\\\vdots&\vdots&&\vdots\\{\widetilde{s}}_{n1}t_1&{\widetilde{s}}_{n2}t_2&\cdots&{\widetilde{s}}_{n9}t_9\\\end{matrix}\right]$


We need this matrix to do the next step, which is to convert to a common scale for all proteins. We do this by normalizing to the amount of protein $\alpha$ in the starting material, which we denote as $a_{\alpha, h}$. In some experiments, if the homogenate is measured directly, this can be calculated from ${{s}_{\alpha ,h}}{{t}_{h}}$.
Alternatively, if a complete set of fractions that entirely represent the homogenate are available, it is preferable to calculate this by summing $s_{\alpha,l}t_l$ over these fractions. In our case, the first six fractions represent a complete set of differential fractions, and thus 

$${a_{\alpha ,h}} = \sum\nolimits_{i = 1}^6 {{s_{\alpha ,i}}{t_i}} $$

Finally, we normalize amounts in any given fraction to amounts in starting material, which we call the relative amount and denote as ${\breve{a}}_{\alpha,l}$



In our example,

$${\overset{\smile}{a}_{\alpha ,l}} = \frac{{{s_{\alpha ,l}}{t_l}}}{{{s_{\alpha ,h}}{t_h}}} = \frac{{{s_{\alpha ,l}}{t_l}}}{{\sum\nolimits_{i = 1}^6 {{s_{\alpha ,i}}{t_i}} }} = \frac{{{{\tilde s}_{\alpha ,l}}{t_l}}}{{\sum\nolimits_{i = 1}^6 {{{\tilde s}_{\alpha ,i}}{t_i}} }}$$


In matrix form, we may write this as 


$\breve{\mathbf{A}}=\mathbf{A}\cdot\mathrm{diag}(1/a_{{\alpha},h})=\left[\begin{matrix}{ {s}}_{11}t_1/a_{1,h}&{ {s}}_{12}t_2/a_{1,h}&\cdots&{ {s}}_{19}t_9/a_{1,h}\\{ {s}}_{21}t_1/a_{2,h}&{ {s}}_{22}t_2/a_{2,h}&\cdots&{ {s}}_{29}t_9/a_{2,h}\\\vdots&\vdots&&\vdots\\{ {s}}_{n1}t_1/a_{n,h}&{ {s}}_{n2}t_2/a_{n,h}&\cdots&{ {s}}_{n9}t_9/a_{n,h}\\\end{matrix}\right]$




 

or simply as  ${\breve{\mathbf{A}}} = (\breve{a}_{\alpha ,l})$.

The function `abundanceTransform` computes this:



```{r, echo=TRUE, eval=TRUE}

AcupMarkers <- relAmtTransform(markerLocR,6, totProt=totProtUse)
round(AcupMarkers, digits=4)
```

The values in `AcupMarkers` represent in principle the relative amount of a given cellular compartment that ends up in a given centrifugation fraction.
 
## Relative Specific Amounts

When examining the distribution of a given protein in different fractions, it is particularly useful to consider its abundance relative to that of total protein.  We refer to this as a Relative Specific Amount (RSA or $r$), which can be calculated as ${r_{\alpha ,l}} = \breve{a}_{\alpha ,l}/ \breve{t}_l$, where the vector ${\breve{\mathbf{t}}} = {\mathbf{t}}/{t_h}$. This describes the fold-enrichment ($r>1$) or depletion ($r<1$) of a protein during the fractionation process, and is analogous to the relative specific activity term used in classical analytical subcellular fractionation.

For a protein $\alpha$ the RSA in fraction $l$ is given by


$r_{\alpha,l}=\frac{\breve{a}}{{\breve{t}}_l}=\frac{s_{\alpha,l}\cdot\sum_{j}^{k}t_i}{\sum_{j}^{k}{s_\alpha}t_i} =\frac{\tilde{s}_{\alpha,l}\cdot\sum_{j}^{k}t_i}{\sum_{j}^{k}{\tilde{s}_\alpha}t_i}$


In matrix form, this is $\mathbf{R}=\left( {{r}_{\alpha ,l}} \right)$.

We can get the RSA matrix from AcupMarkers` and "totProt" as follows:


```{r, echo=TRUE, eval=TRUE}
rsaMarkers <- RSAfromAcup(AcupMarkers, totProt=totProtUse)
round(rsaMarkers, digits=3)
```


Note that we can also obtain the RSA transformed data directly from normalized specific amount data as described in Vignette 2:

```{r, echo=TRUE, eval=TRUE}
markerLocRrsa <- RSAfromS(SS=markerLocR, NstartMaterialFractions=6,
        totProt=totProtUse)

round(markerLocRrsa, digits=3)
```

Finally, if we normalize the RSA matrix so that rows sum to one (via the `apply` function), this yields normalized specific amounts, which are identical to those in `markerLocR`:

```{r, echo=TRUE, eval=TRUE}
round(t(apply(rsaMarkers,1, function(x) x/sum(x))), digits=3)
```


## Simulating proteins resident in multiple subcellular locations

We may simulate data from proteins with multiple residences using the `proteinMix` function. For example, to simulate data from proteins that are distributed in a range of proportions between cytosol and lysosomes, we use this function with normalized amounts of the single-compartment marker profiles:

```{r, echo=TRUE, eval=TRUE}

AcupMarkers <- relAmtTransform(markerLocR, NstartMaterialFractions=6, totProt=totProtUse)
mixCytoLyso <- proteinMix(AcupMarkers, Loc1=1, Loc2=4)
```

The function produces two components.  One component, ‘mix.df’, is a matrix that describes the proportions used to create the simulated multi-resident proteins.  This shows the “true” simulated mixture proportions:

```{r, echo=TRUE, eval=TRUE}
round(mixCytoLyso$mix.df, digits=2)
```

The other component, `mixAmount`,  is a matrix that contains the Acup values (normalized amounts) for the simulated proteins:

```{r, echo=TRUE, eval=TRUE}
round(mixCytoLyso$mixAmount, digits=3)
```

Then we can test the CPA algorithm by first converting this mixture data to RSA's:

```{r, echo=TRUE, eval=TRUE}
mixCytoLysoRSA <- RSAfromAcup(Acup=mixCytoLyso$mixAmount, NstartMaterialFractions=6, totProt=totProtUse)

round(mixCytoLysoRSA, digits=3)

markerLocRrsa <- RSAfromAcup(Acup=AcupMarkers, NstartMaterialFractions=6, 
                             totProt=totProtUse)
```

Finally, we fit the CPA algorithm to this RSA-transformed, simulated data:

```{r, echo=TRUE, eval=TRUE}
mixCytoLysoProp <- proLocAll(protProfileSummary=mixCytoLysoRSA, markerLocR=markerLocRrsa,
                            n.channels=9)
round(mixCytoLysoProp, digits=3)
```


The estimated proportions correspond closely to the proportions used in the simulation.

## Plotting mixtures of proteins with transformations

We may plot the predicted (based on CPA) and expected (based on the simulation parameters) proportional assignments using the `mixturePlot` function. This function evaluates the prediction error by computing the area separating  the predicted and expected protein mixtures via the trapezoidal rule; this is done with the `trapz` function in the `pracma` library, which must have been previously installed. We also need to tell the program which locations were used to generate the mixtures using `Loc1` and `Loc2`:

```{r, echo=TRUE, eval=TRUE}
library(pracma)
mixturePlot(mixProtiProtjProp=mixCytoLysoProp, NstartMaterialFractions=6, Loc1=1, Loc=4,
              mix.df=mixCytoLyso$mix.df, xaxisLab=T, yaxisLab=T)
```

Here we see visually that the estimated proportions match the simulated ones. The area separated by the observed and expected CPA estimates is zero, which is shown in parentheses.

As an alternative, we may obtain CPA estimates using log-transformed relative amounts using `log2Transf=T`:

```{r, echo=TRUE, eval=TRUE}
mixCytoLysoRSAlog2 <- log2(mixCytoLysoRSA + .01)
markerLocRrsalog2 <- log2(markerLocRrsa + .01)

mixCytoLysoPropLog <- proLocAll(protProfileSummary=mixCytoLysoRSAlog2,
                              markerLocR=markerLocRrsalog2,
                            n.channels=9)
round(mixCytoLysoPropLog, digits=3)
mixturePlot(mixProtiProtjProp=mixCytoLysoPropLog, NstartMaterialFractions=6, Loc1=1, Loc=4,
              mix.df=mixCytoLyso$mix.df, xaxisLab=T, yaxisLab=T)

``` 

We see that the estimates are sub-optimal. The area between the observed and expected is 0.268.

## Effect of transformations on error rates for mixtures

Now we consider the effects of fitting CPA to mixtures using a variety of transformations. First, let us consider a mixture of Cyto with each of the other seven compartments in turn, using the ideal RSA-based transformations. We begin by setting up the plot area for a 4 by 2 array of plots. Optionally, to control the size of the window, we may want to explicitly open a window using `windows(height=10, width=7)`. Next we fix one component of the mixture to the first, which is Cyto, and loop over the other 7 subcellular compartments, creating mixtures of the AcupMarkers values. For each case, we transform these mixtures to RSA and obtain CPA mixing proportion estimates from these RSA-transformed mixtures and compute the area-based prediction errors. 

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}
par(mfrow=c(4,2))
i <- 1
mixErrorMat <- NULL
for (j in 2:8) {   
   # Create the mixture of Cyto (i = 1) with compartment j
   mixProtiProtj <- proteinMix(Acup=AcupMarkers, Loc1=i, Loc2=j)

   # Tranform the mixtures to relative specific amounts
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtj$mixAmount, NstartMaterialFractions=6,  
                     totProt=totProtUse)
    
   # Find the constrained proportional assigments (CPA)                 
   mixProtiProtjProp <- proLocAll(protProfileSummary=mixProtiProtjRSA,
                        markerLocR=markerLocRrsa, n.channels=9)
   
   # Plot the results, including the area-based error estimate, 
   #    and collect the area-based errors (errorReturn=T)                         
    mixResult <- mixturePlot(mixProtiProtjProp=mixProtiProtjProp, NstartMaterialFractions=6, 
                             Loc1=i, Loc2=j, mix.df=mixProtiProtj$mix.df, errorReturn = T)
    mixErrorMat <- rbind(mixErrorMat, mixResult)         
}
```

All seven mixtures have essentially zero area-based error, as we expect. We can examine these errors with more precision as follows:

```{r, echo=TRUE, eval=TRUE}
mixErrorMat
```

Next, consider a log2 transformation `y = log2(prop + 0.01)` of the RSA values before carrying out the CPA procedure. Aside from this log2 transformation, the other R commands are the same as above.


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}
par(mfrow=c(4,2))  # create a 4 by 2 matrix of plots

mixErrorMat <- NULL
i=1  # fix one component of the mixture to the first, which is Cyto
for (j in 2:8) {   # mix with each of the other compartments, and plot
   #j=4
   # Create mixture of subcellular locations i and j using "Acupmarkers"
   mixProtiProtj <- proteinMix(Acup=AcupMarkers, Loc1=i, Loc2=j)
 
   # Transform these mixtures to RSA
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtj$mixAmount, NstartMaterialFractions=6,  
                     totProt=totProtUse)
   
   # Take a log2 transformation
   log2MixProtiProtjRSA <- log2(mixProtiProtjRSA + 0.01)
   log2MarkerLocRrsa <- log2(markerLocRrsa + 0.01)
                     
   # Obtain CPA estimates from the RSA-transformed mixtures
   mixProtiProtjProp <- proLocAll(protProfileSummary=log2MixProtiProtjRSA,
                        markerLocR=log2MarkerLocRrsa, n.channels=9)
                            
   # Plot the results, including the area-based error estimate
   mixResult <- mixturePlot(mixProtiProtjProp=mixProtiProtjProp, NstartMaterialFractions=6, 
                            Loc1=i, Loc2=j, mix.df=mixProtiProtj$mix.df, errorReturn = T)
   mixErrorMat <- rbind(mixErrorMat, mixResult)
             
}
```

The area-based errors are as follows:

```{r, echo=TRUE, eval=TRUE}
mixErrorMat
```


Now consider using normalized specific amounts, which is what we would use for these mixtures if the amounts of starting material in each fraction were not available. The first two function calls are as above, but we obtain these normalized specific values by constraining the RSA values to sum to 1 before using `proLocAll` with `markerLocRuse' to compute the CPA estimates.


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}

#windows(height=10, width=7)
par(mfrow=c(4,2))  # create a 4 by 2 matrix of plots

mixErrorMat <- NULL
i=1  # fix one component of the mixture to the first, which is Cyto
for (j in 2:8) {   # mix with each of the other compartments, and plot
   #j=4
   # Create mixture of subcellular locations i and j using "Acupmarkers"
   mixProtiProtj <- proteinMix(Acup=AcupMarkers, Loc1=i, Loc2=j)
 
   # Transform these mixtures to RSA
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtj$mixAmount, NstartMaterialFractions=6,  
                     totProt=totProtUse)
   
   # Now convert to normalized specific amounts
   mixProtiProtjSpecific <- t(apply(mixProtiProtjRSA,1, function(x) x/sum(x)))
   markerLocRspecific <- t(apply(markerLocRrsa,1, function(x) x/sum(x)))
                     
   # Obtain CPA estimates from the RSA-transformed mixtures
   mixProtiProtjProp <- proLocAll(protProfileSummary=mixProtiProtjSpecific,
                        markerLocR=markerLocRspecific, n.channels=9)
                            
   # Plot the results, including the area-based error estimate, 
   #    and collect the area-based errors (errorReturn=T)
   mixResult <- mixturePlot(mixProtiProtjProp=mixProtiProtjProp, NstartMaterialFractions=6, 
                            Loc1=i, Loc2=j, mix.df=mixProtiProtj$mix.df, errorReturn = T)
   mixErrorMat <- rbind(mixErrorMat, mixResult)
             
}
```

Here are the area-based errors:

```{r, echo=TRUE, eval=TRUE}
mixErrorMat
```

Finally we consider log2-transformed values of normalized specific amounts. This set of plots is similar to the last one except a transformation `y = log2(prop + 0.01)` is applied before obtaining CPA estimates.


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}

#windows(height=10, width=7)
par(mfrow=c(4,2))  # create a 4 by 2 matrix of plots

mixErrorMat <- NULL
i=1  # fix one component of the mixture to the first, which is Cyto
for (j in 2:8) {   # mix with each of the other compartments, and plot
   #j=4
   # Create mixture of subcellular locations i and j using "Acupmarkers"
   mixProtiProtj <- proteinMix(Acup=AcupMarkers, Loc1=i, Loc2=j)
 
   # Transform these mixtures to RSA
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtj$mixAmount, NstartMaterialFractions=6,  
                     totProt=totProtUse)
   
   # Now convert to normalized specific amounts
   mixProtiProtjSpecific <- t(apply(mixProtiProtjRSA,1, function(x) x/sum(x)))
   markerLocRspecific <- t(apply(markerLocRrsa,1, function(x) x/sum(x)))
   log2MixProtiProtjSpecific <- log2(mixProtiProtjSpecific + 0.01)
   log2MarkerLocRspecific <- log2(markerLocRspecific + 0.01)
                     
   # Obtain CPA estimates from the RSA-transformed mixtures
   mixProtiProtjProp <- proLocAll(protProfileSummary=log2MixProtiProtjSpecific,
                        markerLocR=log2MarkerLocRspecific, n.channels=9)
                            
   # Plot the results, including the area-based error estimate
    mixResult <- mixturePlot(mixProtiProtjProp=mixProtiProtjProp, NstartMaterialFractions=6, 
                             Loc1=i, Loc2=j, mix.df=mixProtiProtj$mix.df, errorReturn = T)
    mixErrorMat <- rbind(mixErrorMat, mixResult)         
}
```

Here are the area-based errors:

```{r, echo=TRUE, eval=TRUE}
mixErrorMat
```

We may make these plots for all 8*7/2 = 28 possible pairs of mixtures. The function `mixturePlotPanel` does this for either RSA-transformed mixtures (`fitType="rsa"`) or mixtures that have been further transformed to normalized specific amounts (`fitType="specAmt"`). An option is also available to transform either of these using `y = log2(x + eps)', where `eps` is a small number; by default, `eps = 0.01`. The option `errorReturn = T` causes the function to return a table of all area-based errors. By summing the third column, we obtain a global measure of error.

The following code will produce all four combinations of fit type (rsa or specfAmt) and log transformation (True or False). The plots are written as pdf files to accommodate the large number of plots.

First is rsa without and with log-transformed values:

``` 
setwd("c:\\temp5")

fitType <- "rsa"
log2Transf <- F

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtUse, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])


fitType <- "rsa"
log2Transf <- T

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtUse, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])

```

Next are using normalized specific amounts, without and with log-transformed values:

``` 
setwd("c:\\temp5")

fitType <- "specAmt"
log2Transf <- F

pdf(file=paste("CPA assignProts Jadot", fitType=fitType, transf=log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtUse, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])


fitType <- "specAmt"
log2Transf <- T

pdf(file=paste("CPA assignProts Jadot", fitType, log2Transf, ".pdf"), width=7, height=11)

errorAll <- mixturePlotPanel(Acup=AcupMarkers, totProt=totProtUse, errorReturn = T, 
              fitType=fitType, log2Transf=log2Transf)
dev.off()

# overall errors are as follows:
errorAll
sum(errorAll[,3])

```







