---
title: 'Vignette 3: Data transformations: Notation and mixtures using protlocassign'
author: "Dirk Moore"
date: "`r Sys.Date()`"
output:
  #html_document:
  #  toc: yes
  #  df_print: paged
  word_document:
    fig_caption: yes
    toc: yes
  #pdf_document:
  #  toc: yes
vignette: |
  %\VignetteIndexEntry{protlocassign} %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff = 150
		),
	options(width = 150),
	eval = TRUE
)
```

## Introduction

As explained in the main text and in Vignette 2, there are different ways to transform protein profile data.  Here, we describe a way to explore the effect of using transformed data with CPA.  Briefly, we conduct different data transformations on a set of theoretical proteins that have a range of distributions between two cellular compartments, and then conduct CPA and determine how well it predicts the original distribution.  For our simulations, to create the theoretical proteins, we use data from the experiment from Tannous et al which consists of a TMT-MS2 analysis of six differential fraction (N, M, L1, L2, P, and S) obtained from centrifugation of a rat liver homogenate and three fractions from a Nycodenz density gradient separation of the differential fraction L1 (Nyc1, Nyc2, and Nyc3).

In our procedure, we first use the eight compartment profiles generated from the marker protein set to simulate a set of eight theoretical proteins that wholly reside in each of the respective compartments.  We then mix these in defined amounts to simulate proteins that are distributed in varying proportions between two compartments. Note that data from mass spectrometry experiments represent specific amounts ${s_{\alpha ,l}}$ of a protein $\alpha$ in fraction $l$, with the same amount of total protein being analyzed for each sample (fraction). For conducting our simulations, we first must transform this data into relative amounts, so that each protein has precisely the same total amount in the initial starting material used for fractionation.

## Computation of Relative Amounts

Consider an $n$ by 9 matrix of specific amounts that details the average distribution of $n$ proteins (or other species of interest) among 9 fractions:
$$
\mathbf{S}=\left[ \begin{matrix}
   {{s}_{11}} & {{s}_{12}} & \cdots  & {{s}_{19}}  \\
   {{s}_{21}} & {{s}_{22}} & \cdots  & {{s}_{29}}  \\
   \vdots  & \vdots  & {} & \vdots   \\
   {{s}_{n1}} & {{s}_{n2}} & \cdots  & {{s}_{n9}}  \\
\end{matrix} \right]
$$

Each row $\alpha$ represents a mean profile for a protein or other species of interest $\alpha$, with each profile consisting of $f=9$ fractions. 

In the Vignette 1 we actually used a normalized specific amount (NSA), denoted here as $\tilde{s}_{\alpha,l}$  calculated as follows:


$$ {{\tilde{s}}_{\alpha ,l}}=\frac{{{s}_{\alpha ,l}}}{\sum\limits_{j=1}^{f}{{{s}_{\alpha ,j}}}} $$
In some experiments, where bookkeeping is not practicable, these are the only values available for using the CPA procedure. However, with appropriate bookkeeping, one can estimate the amounts of total protein present in different samples obtained from a set amount of starting material. These include the total protein content of the starting material (designated $t_h$) and the total protein content of any given fraction (designated $t_l$ for fraction $l$). We can use these to calculate the amount of protein in arbitrary units in fraction $l$ derived from a set amount of starting material as $a_l = s_l t_l$. Note that as $a_l$ is in arbitrary units one can do the same calculation using either $s_l$ or ${\tilde s_l}$, as later, these will yield the same values when calculating relative amounts (see below).

To see how this works in `protlocassign`, let us consider the Jadot reference protein profiles which we generate using the `locationProfileSetup` function. As in earlier vignettes, we rename `protNSA_AT5tmtMS2` as `protNSA` and we tailor the output using the `round` function.

```{r, echo=TRUE, eval=TRUE}
library(protlocassign)
protNSA <- protNSA_AT5tmtMS2
refLocationProfilesNSA <- locationProfileSetup(profile=protNSA, 
                           markerList=markerListJadot, numDataCols=9)
round(refLocationProfilesNSA, digits=3)

```


Here, each row represents the profile $\tilde{s}_l$ for a protein resident solely in a particular cellular compartment. The amount of total protein derived from a set amount of starting material from all fractions in the experiment used for these vignettes is in `totProtAT5`, a vector supplied with the `protassign` package. For convenience we rename it `totProt`.

```{r, echo=TRUE, eval=TRUE}
totProt <- totProtAT5
round(totProt, digits=3)
```

We denote these values using a vector $\mathbf{t}= ({t_1},{t_2}, \ldots ,{t_9})$.

As noted above, for protein $\alpha$ in fraction $l$, we have ${{a}_{\alpha ,l}}={{\tilde{s}}_{\alpha ,l}}{{t}_{l}}$. In matrix form,


$\mathbf{A}=\widetilde{\mathbf{S}}\cdot\mathrm{diag}(\mathbf{t})=\left[\begin{matrix}{\widetilde{s}}_{11}t_1&{\widetilde{s}}_{12}t_2&\cdots&{\widetilde{s}}_{19}t_9\\{\widetilde{s}}_{21}t_1&{\widetilde{s}}_{22}t_2&\cdots&{\widetilde{s}}_{29}t_9\\\vdots&\vdots&&\vdots\\{\widetilde{s}}_{n1}t_1&{\widetilde{s}}_{n2}t_2&\cdots&{\widetilde{s}}_{n9}t_9\\\end{matrix}\right]$


We need this matrix to do the next step, which is to convert to a common scale for all proteins. We do this by normalizing to the amount of protein $\alpha$ in the starting material, which we denote as $a_{\alpha, h}$. In some experiments, if the homogenate is measured directly, this can be calculated from ${\widetilde{s}_{\alpha ,h}}{{t}_{h}}$.
Alternatively, if a complete set of fractions that entirely represent the homogenate are available, it is preferable to calculate this by summing $\widetilde{s}_{\alpha,l}t_l$ over these fractions. In our case, the first six fractions (N, M, L1, L2, P, and  S) are a complete set of differential fractions that represent the starting material, and thus 

$${a_{\alpha ,h}} = \sum\nolimits_{i = 1}^6 {{\tilde{s}_{\alpha ,i}}{t_i}} $$
and $t_h=\sum_{i=1}^{6}t_i$. Note that this is readily calculated by summing the first six elements of $\mathbf{t}$.

```{r, echo=TRUE, eval=TRUE}
sum(totProt[1:6])
```

(The last three "Nyc" columns were derived from L1, so we do not include them in the sum.)

Finally, we normalize amounts in any given fraction to amounts in starting material, which we call the relative amount, designamted here as `Acup`, and denote as ${\breve{a}}_{\alpha,l}$



In our example,

$${\overset{\smile}{a}_{\alpha ,l}} =\frac{a_l}{a_h} = \frac{{{s_{\alpha ,l}}{t_l}}}{{{s_{\alpha ,h}}{t_h}}} = \frac{{{s_{\alpha ,l}}{t_l}}}{{\sum\nolimits_{i = 1}^6 {{s_{\alpha ,i}}{t_i}} }} = \frac{{{{\tilde s}_{\alpha ,l}}{t_l}}}{{\sum\nolimits_{i = 1}^6 {{{\tilde s}_{\alpha ,i}}{t_i}} }}$$


In matrix form, we may write this as 


$\breve{\mathbf{A}}=\mathbf{A}\cdot\mathrm{diag}(1/a_{{\alpha},h})=\left[\begin{matrix}{ \tilde{s}}_{11}t_1/a_{1,h}&{ \tilde{s}}_{12}t_2/a_{1,h}&\cdots&{ \tilde{s}}_{19}t_9/a_{1,h}\\{ \tilde{s}}_{21}t_1/a_{2,h}&{ \tilde{s}}_{22}t_2/a_{2,h}&\cdots&{ \tilde{s}}_{29}t_9/a_{2,h}\\\vdots&\vdots&&\vdots\\{ \tilde{s}}_{n1}t_1/a_{n,h}&{ \tilde{s}}_{n2}t_2/a_{n,h}&\cdots&{ \tilde{s}}_{n9}t_9/a_{n,h}\\\end{matrix}\right]$

 
or simply as  ${\breve{\mathbf{A}}} = [\breve{a}_{\alpha ,l}]$.

The function `AcupFromNSA` computes this:

```{r, echo=TRUE, eval=TRUE}

refLocationProfilesAcup <- AcupFromNSA(NSA=refLocationProfilesNSA, NstartMaterialFractions=6, 
                               totProt=totProt)
round(refLocationProfilesAcup, digits=5)
```

The values in `refLocationProfilesAcup` represent in principle the relative amount of a given cellular compartment that ends up in a given centrifugation fraction.
 
## Computation of Relative Specific Amounts from Relative Amounts

When examining the distribution of a given protein in different fractions, it is particularly useful to consider its abundance relative to that of total protein.  We refer to this as a Relative Specific Amount (RSA or $r$), which can be calculated as ${r_{\alpha ,l}} = \breve{a}_{\alpha ,l}/ \breve{t}_l$, where the vector ${\breve{\mathbf{t}}} = {\mathbf{t}}/{t_h}$. 
For a protein $\alpha$ the RSA in fraction $l$ is given by


$r_{\alpha,l}=\frac{\breve{a}}{{\breve{t}}_l}=\frac{s_{\alpha,l}\cdot\sum_{j}^{k}t_i}{\sum_{j}^{k}{s_\alpha}t_i} =\frac{\tilde{s}_{\alpha,l}\cdot\sum_{j}^{k}t_i}{\sum_{j}^{k}{\tilde{s}_\alpha}t_i}$


In matrix form, this is $\mathbf{R}=\left[ {{r}_{\alpha ,l}} \right]$.

We can get the RSA matrix from `refLocationProfilesAcup` and `totProt` as follows:


```{r, echo=TRUE, eval=TRUE}
refLocationProfilesRSA <- RSAfromAcup(refLocationProfilesAcup, totProt=totProt)
round(refLocationProfilesRSA, digits=3)
```


Note that we can also obtain the RSA transformed data directly from normalized specific amount data as described in Vignette 2:

```{r, echo=TRUE, eval=TRUE}
refLocationProfilesRSA_2 <- RSAfromNSA(NSA=refLocationProfilesNSA,
                                NstartMaterialFractions=6, totProt=totProt)
```

This yields precisely the same values as calculated previously using `locationProfileSetup`.

```{r, echo=TRUE, eval=TRUE}
identical(refLocationProfilesRSA, refLocationProfilesRSA_2)
```

Finally, if we normalize an RSA profile so that the rows sum to one, this yields a normalized specific amount profile. Consider example the matrix `refLocationProfilesRSA`, which contains the RSA-transformed compartment profiles. We normalize the rows using the `apply` function, and then transpose using the `t` function to yield a matrix of the normalized specific amounts; these values are essentially identical to those that we started with in `refLocationProfiles`:

```{r, echo=TRUE, eval=TRUE}
refLocationProfilesNSA_2 <- t(apply(
                                    refLocationProfilesRSA_2, 1,
                                   function(x) x/sum(x)))

```

Note that this is not identical to the values obtained previously using the `locationProfileSetup` function with the protein profiles containing NSA data because of internal precision issues, but is essentially equivalent.

```{r, echo=TRUE, eval=TRUE}
identical(refLocationProfilesNSA_2, as.matrix(refLocationProfilesNSA))  
all.equal(refLocationProfilesNSA_2, as.matrix(refLocationProfilesNSA), tolerance=0)
```


## Simulating proteins resident in multiple subcellular locations

We may simulate data from proteins with multiple residences using the `proteinMix` function. For example, to simulate data from proteins that are distributed in a range of proportions between cytosol and lysosomes, we use this function with relative amounts of their single-compartment profiles. We specify the mixing locations by the arguments `Loc1=1` and `Loc2=4`, which are the row numbers of the desired compartment profiles in `refLocationProfiles`:

```{r, echo=TRUE, eval=TRUE}

refLocationProfilesAcup <- AcupFromNSA(NSA=refLocationProfilesNSA, NstartMaterialFractions=6, 
                               totProt=totProt)
data.frame(rownames(refLocationProfilesAcup))
mixCytoLysoAcup <- proteinMix(AcupRef=refLocationProfilesAcup, Loc1=1, Loc2=4)
```

The function produces two components.  One component, ‘input.prop’, is a matrix that describes the proportions used to create the simulated multi-resident proteins.  This shows the “true” simulated mixture proportions:

```{r, echo=TRUE, eval=TRUE}
round(mixCytoLysoAcup$input.prop, digits=2)
```

The other component, `Acup`,  is a matrix that contains the Acup values (relative amounts) for the simulated proteins:

```{r, echo=TRUE, eval=TRUE}
round(mixCytoLysoAcup$Acup, digits=3)
```

Then we can test the CPA algorithm by first converting this mixture data to RSA's:

```{r, echo=TRUE, eval=TRUE}
mixCytoLysoRSA <- RSAfromAcup(Acup=mixCytoLysoAcup$Acup, 
                              NstartMaterialFractions=6, totProt=totProt)

round(mixCytoLysoRSA, digits=3)

```

Finally, we fit the CPA algorithm to this RSA-transformed, simulated data using the previously generated RSA-transformed marker protein profiles:

```{r, echo=TRUE, eval=TRUE}
mixCytoLysoCPAfromRSA <- fitCPA(profile=mixCytoLysoRSA,
                             refLocationProfiles=refLocationProfilesRSA,
                            numDataCols=9)
round(mixCytoLysoCPAfromRSA, digits=3)
```


The estimated proportions correspond closely to the proportions used in the simulation.

Plotting these estimates against the “true” distributions of the simulated proteins provides insights into different transformations on the goodness of fit. We first illustrate this using simulated proteins distributed in varying amounts between the cytoplasm and the lysosome. We then extend this to simulated proteins distributed cytoplasm and each of the other compartments.  Finally, we provide a function that performs simulations on proteins residing between all pairs of compartments.

## Plotting mixtures of proteins with transformations

We can use the `mixturePlot` function to evaluate the effect of different data transformations used to represent protein profiles on the compartmental distributions estimated by CPA. The function produces graphical representations by plotting the theoretical distribution (based on simulation parameters, x-coordinate) versus the predicted values (based on CPA, y-coordinate).  
This function also evaluates the prediction error by computing the area separating the predicted and expected protein mixtures via the trapezoidal rule; this is done with the `trapz` function in the `pracma` library, which must have been previously installed. We also need to tell the program which two locations were used to generate the mixtures using `Loc1` and `Loc2`. As our first example, we examine the results CPA using the RSA transformation on simulated proteins distributed between Cyto and Lyso.  

```{r, echo=TRUE, eval=TRUE}
library(pracma)
par(mfrow=c(1,1))  # reset window for a single plot
mixturePlot(mixProtiProtjCPA=mixCytoLysoCPAfromRSA, 
            NstartMaterialFractions=6, Loc1=1, Loc2=4,
            input.prop=mixCytoLysoAcup$input.prop, xaxisLab=T, yaxisLab=T)
```

Here we see visually that the estimated proportions match the simulated ones. The area separated by the observed and expected CPA estimates, an indication of the goodness of fit, is nearly zero, and is shown in parentheses.

As an alternative, we may obtain CPA estimates using log-transformed relative amounts, first adding a small constant to avoid taking the log of zero:

```{r, echo=TRUE, eval=TRUE}
mixCytoLysoRSAlog2 <- log2(mixCytoLysoRSA + 0.001)
refLocationProfilesRSAlog2 <- log2(refLocationProfilesRSA + 0.001)

mixCytoLysoCPAfromRSALog2 <- fitCPA(profile=mixCytoLysoRSAlog2,
                              refLocationProfiles=refLocationProfilesRSAlog2,
                            numDataCols=9)
round(mixCytoLysoCPAfromRSALog2, digits=3)
mixturePlot(mixProtiProtjCPA=mixCytoLysoCPAfromRSALog2, 
              NstartMaterialFractions=6, Loc1=1, Loc=4,
              input.prop=mixCytoLysoAcup$input.prop, xaxisLab=T, yaxisLab=T)

``` 

We see that the estimates are sub-optimal. The area between the observed and expected is 0.285.

## Effect of transformations on error rates for mixtures

Now we consider the effects of fitting CPA to mixtures using a variety of transformations. First, let us consider a mixture of Cyto with each of the other seven compartments in turn, using the ideal RSA-based transformations. We begin by setting up the plot area for a 4 by 2 array of plots. Optionally, to control the size of the window, we may want to explicitly open a window using `windows(height=10, width=7)`. Next we fix one component of the mixture to the first, which is Cyto, and loop over the other 7 subcellular compartments, creating mixtures of the `refLocationProfilesAcup` values. For each case, we transform these mixtures to RSA and obtain CPA mixing proportion estimates from these RSA-transformed mixtures and compute the area-based prediction errors. 
These values are stored in a dataframe `mixErrorMat` which is then renamed to avoid overwriting since multiple mixtures and transformations are being explored.

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}
par(mfrow=c(4,2))
i <- 1
mixErrorMat <- NULL
for (j in 2:8) {   
   # Create the mixture of Cyto (i = 1) with compartment j
   mixProtiProtjAcup <- proteinMix(Acup=refLocationProfilesAcup, Loc1=i, Loc2=j)

   # Tranform the mixtures to relative specific amounts
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtjAcup$Acup, 
                        NstartMaterialFractions=6, totProt=totProt)
    
   # Find the constrained proportional assigments (CPA)                 
   mixProtiProtjCPAfromRSA <- fitCPA(profile=mixProtiProtjRSA,
                    refLocationProfiles=refLocationProfilesRSA, numDataCols=9)
   
   # Plot the results, including the area-based error estimate, 
   #    and collect the area-based errors (errorReturn=T)                         
    mixResult <- mixturePlot(mixProtiProtjCPA=mixProtiProtjCPAfromRSA, 
                             NstartMaterialFractions=6, Loc1=i, Loc2=j, 
                             input.prop=mixProtiProtjAcup$input.prop, errorReturn = T)
    mixErrorMat <- rbind(mixErrorMat, mixResult)         
}
mixErrorAllCytoRSA <- mixErrorMat
```

All seven mixtures have essentially zero area-based error, as we expect. We can examine these errors with more precision as follows:

```{r, echo=TRUE, eval=TRUE}
mixErrorAllCytoRSA
```

Next, consider a log2 transformation `y = log2(prop + 0.001)` of the RSA values before carrying out the CPA procedure. Aside from this log2 transformation, the other R commands are the same as above.


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}
par(mfrow=c(4,2))  # create a 4 by 2 matrix of plots

mixErrorMat <- NULL
refLocationProfilesRSAlog2 <- log2(refLocationProfilesRSA + 0.001)  # transform marker profiles
i=1  # fix one component of the mixture to the first, which is Cyto
for (j in 2:8) {   # mix with each of the other compartments, and plot
   #j=4
   # Create mixture of subcellular locations i and j using "refLocationProfilesAcup"
   mixProtiProtj <- proteinMix(Acup=refLocationProfilesAcup, Loc1=i, Loc2=j)
 
   # Transform these mixtures to RSA
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtj$Acup, NstartMaterialFractions=6,  
                     totProt=totProt)
   
   # Take a log2 transformation
   MixProtiProtjRSAlog2 <- log2(mixProtiProtjRSA + 0.001)
   
                     
   # Obtain CPA estimates from the log2 RSA-transformed mixtures
   mixProtiProtjCPAfromRSAlog2 <- fitCPA(profile=MixProtiProtjRSAlog2,
                        refLocationProfiles=refLocationProfilesRSAlog2, numDataCols=9)
                            
   # Plot the results, including the area-based error estimate
   mixResult <- mixturePlot(mixProtiProtjCPA=mixProtiProtjCPAfromRSAlog2, 
                            NstartMaterialFractions=6, Loc1=i, Loc2=j, 
                            input.prop=mixProtiProtj$input.prop, errorReturn = T)
   mixErrorMat <- rbind(mixErrorMat, mixResult)
             
}
mixErrorAllCytoRSAlog2 <- mixErrorMat
```

The area-based errors are as follows:

```{r, echo=TRUE, eval=TRUE}
mixErrorAllCytoRSAlog2
```


Now consider using normalized specific amounts, which is what we would use for these mixtures if the amounts of starting material in each fraction were not available. The first two function calls are as above, but we obtain these normalized specific amounts by constraining the RSA values to sum to 1 before using `fitCPA` with `refLocationProfilesNSA' to compute the CPA estimates.


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}
refLocationProfilesNSA <- t(apply(refLocationProfilesRSA, 1, function(x) x/sum(x)))
#windows(height=10, width=7)
par(mfrow=c(4,2))  # create a 4 by 2 matrix of plots

mixErrorMat <- NULL
i=1  # fix one component of the mixture to the first, which is Cyto
for (j in 2:8) {   # mix with each of the other compartments, and plot
   #j=4
   # Create mixture of subcellular locations i and j using "refLocationProfilesAcup"
   mixProtiProtjAcup <- proteinMix(Acup=refLocationProfilesAcup, Loc1=i, Loc2=j)
 
   # Transform these mixtures to RSA
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtjAcup$Acup, 
                        NstartMaterialFractions=6, totProt=totProt)
   
   # Now convert to normalized specific amounts
   mixProtiProtjNSA <- t(apply(mixProtiProtjRSA,1, function(x) x/sum(x)))
                     
   # Obtain CPA estimates from the normalized specific amount-transformed mixtures
   mixProtiProtjCPAfromNSA <- fitCPA(profile=mixProtiProtjNSA,
                    refLocationProfiles=refLocationProfilesNSA, numDataCols=9)
                            
   # Plot the results, including the area-based error estimate, 
   #    and collect the area-based errors (errorReturn=T)
   mixResult <- mixturePlot(mixProtiProtjCPA=mixProtiProtjCPAfromNSA, 
                            NstartMaterialFractions=6, Loc1=i, Loc2=j, 
                            input.prop=mixProtiProtjAcup$input.prop, errorReturn = T)
   mixErrorMat <- rbind(mixErrorMat, mixResult)
             
}
mixErrorAllCytoNSA <- mixErrorMat

```

Here are the area-based errors:

```{r, echo=TRUE, eval=TRUE}
mixErrorAllCytoNSA
```

Finally, we consider log2-transformed values of normalized specific amounts. This set of plots is similar to the last one except a transformation `y = log2(prop + 0.001)` is applied before obtaining CPA estimates. We calculate `refLocationProfilesuse` by constraining the row totals to be 1. 


```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height = 10}
refLocationProfilesNSA <- t(apply(refLocationProfilesRSA, 1, function(x) x/sum(x)))
#windows(height=10, width=7)
par(mfrow=c(4,2))  # create a 4 by 2 matrix of plots

mixErrorMat <- NULL
refLocationProfilesNSAlog2 <- log2(refLocationProfilesNSA + 0.001) # do this here
i=1  # fix one component of the mixture to the first, which is Cyto
for (j in 2:8) {   # mix with each of the other compartments, and plot
   #j=4
   # Create mixture of subcellular locations i and j using "Acupmarkers"
   mixProtiProtjAcup <- proteinMix(Acup=refLocationProfilesAcup, Loc1=i, Loc2=j)
 
   # Transform these mixtures to RSA
   mixProtiProtjRSA <- RSAfromAcup(Acup=mixProtiProtjAcup$Acup, 
                          NstartMaterialFractions=6,  totProt=totProt)
   
   # Now convert to normalized specific amounts
   mixProtiProtjNSA <- t(apply(mixProtiProtjRSA,1, function(x) x/sum(x)))
   MixProtiProtjNSAlog2 <- log2(mixProtiProtjNSA + 0.001)
                     
   # Obtain CPA estimates from the RSA-transformed mixtures
   mixProtiProtjCPAfromNSAlog2 <- fitCPA(profile=MixProtiProtjNSAlog2,
                        refLocationProfiles=refLocationProfilesNSAlog2, numDataCols=9)
                            
   # Plot the results, including the area-based error estimate
    mixResult <- mixturePlot(mixProtiProtjCPA=mixProtiProtjCPAfromNSAlog2, 
                             NstartMaterialFractions=6, Loc1=i, Loc2=j, 
                             input.prop=mixProtiProtjAcup$input.prop, errorReturn = T)
    mixErrorMat <- rbind(mixErrorMat, mixResult)         }
mixErrorAllCytoNSAlog2 <- mixErrorMat
```

Here are the area-based errors:

```{r, echo=TRUE, eval=TRUE}
mixErrorAllCytoNSAlog2
```

## References

Jadot M, Boonen M, Thirion J, Wang N, Xing J, Zhao C, Tannous A, Qian M, Zheng H, Everett JK, Moore DF, Sleat DE, Lobel P (2016) Accounting for protein subcellular localization: a compartmental map of the rat liver proteome. Molecular and Cellular Proteomics 16, 194-212. doi:10.1074/mcp.M116.064527  PMCID: PMC5294208

Tannous A, Boonen M, Zheng H, Zhao C, Germain C, Moore D, Sleat D, Jadot M, Lobel P. (2020) Comparative Analysis of Quantitative Mass Spectrometric Methods for Subcellular Proteomics. Journal of Proteome Research. Journal of Proteome Research 19, 1718-1730. doi: 10.1021/acs.jproteome.9b00862
