---
title: "Vignette 5: Distance, outlier detection, and weighted mean profile calculation"
author: "Dirk Moore"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output:
  word_document:
  #  html_document:
  #mathjax:  default
  fig_caption:  true
toc: true
#section_numbering: true
#css: ggsci.css
#output:
#  word_document:
#    reference_docx: word-styles-reference-03.docx
#output: pdf_document
#fontsize: 10pt
vignette: >
  %\VignetteIndexEntry{protlocassign}
  %\VignetteEngine{knitr::rmarkdown}
---

  ```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(
    keep.blank.line = TRUE,
    width.cutoff = 150
  ),
  options(width = 150),
  eval = TRUE
)
```



## Nearest proteins in a data set
We may find the proteins with profiles nearest to a given protein using the function "nearestProts". Distance is computed as the Euclidean distance between profiles. To use the function, we first use the R function `dist` to create a distance matrix for the proteins in a list of mean profiles, such as `protProfileNSA_AT5tmtMS2`. 

```{r, echo=TRUE, eval=TRUE}
library(protlocassign)
distUse <- dist(protProfileNSA_AT5tmtMS2[,1:9], method="euclidean")
```
Then select the protein names:

```{r, echo=TRUE, eval=TRUE}
protsUse <- rownames(protProfileNSA_AT5tmtMS2)
```
Finally, provide a protein name. Here, for the protein "CTSD", we find the 10 nearest proteins.

```{r, echo=TRUE, eval=TRUE}
nearestProts(protName="CTSD", n.nearest=10,  distProts=distUse, protNames=protsUse,
             profile=protProfileNSA_AT5tmtMS2)
```

Instead of using normalized specific amounts, we may transform them to relative specific amounts:

```{r, echo=TRUE, eval=TRUE}
protProfileLevelsRSA <- RSAfromNSA(NSA=protProfileNSA_AT5tmtMS2[,1:9],
                                 NstartMaterialFractions=6, totProt=totProtAT5)
distUseRSA <- dist(protProfileLevelsRSA, method="euclidean")
nearestProts(protName="CTSD", n.nearest=10,  distProts=distUseRSA, protNames=protsUse,
             profile=protProfileLevelsRSA)

```

## References

Jadot M, Boonen M, Thirion J, Wang N, Xing J, Zhao C, Tannous A, Qian M, Zheng H, Everett JK, Moore DF, Sleat DE, Lobel P (2016) Accounting for protein subcellular localization: a compartmental map of the rat liver proteome. Molecular and Cellular Proteomics 16, 194-212. doi:10.1074/mcp.M116.064527  PMCID: PMC5294208

Tannous A, Boonen M, Zheng H, Zhao C, Germain C, Moore D, Sleat D, Jadot M, Lobel P. (2020) Comparative Analysis of Quantitative Mass Spectrometric Methods for Subcellular Proteomics. Journal of Proteome Research. Journal of Proteome Research 19, 1718-1730. doi: 10.1021/acs.jproteome.9b00862

